Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-03-01 VegOzone.tex
Restructure introduction in accordance to Lisa's comments - not spell checked.

\Author[1]{Stefanie}{Falk}
\Author[2]{Ane Victoria}{Vollsnes}
\Author[2]{Aud Else}{Berglen Erikson}
\Author[3]{Lisa}{Emberson}
\Author[3]{Connie}{O'Neill}
\Author[1]{Frode}{Stordal}
\Author[1]{Terje}{Koren Berntsen}

@@

\begin{itemize}
\item Wealth of evidence demonstrating ozone damage to vegetation (perhaps focus on key evidence from northern climates and key species (e.g. birch, Norway srpuce etc... - this might help to make it clear why we focus on these species with the PODy modelling))
\item Although O3 trends stagnant or decreasing in Europe there is unique photochemcistry over the arctic (causing spring time peaks in ozone) that may mean ozone damage continues to be an issue
\item The earlier timing of growing seasons with climate change and 24 hour sunlight may also enhance the exposure of plants during those times when they are physiologically active to higher [O3]
\item Research also clearly shows how ecosystem sensitivity to vegetation is influenced by climate variables, both because this can influence ozone uptake (dose) as well as the impact of altered ozone doses (can add in some relevant detail here on ozone damage mechanisms - again perhaps with a focus on what we know for colder climates)
\item Risk assessment methods have been developed that allow for ozone uptake (i.e. PODy)
\item Comparison of these methods with previously used concentration based methods (e.g. AOT40) show that PODy is better able to represent the risk to vegetation from ozone, especially in the more climate extreme parts of Europe (see Mills et al 2011 paper attached - that also includes some work for clover that we could use)
\item These methods rely on accurate representation of plant physiological response to climate variables (i.e. the species or plant funtional type specific responses used to parameterise risk assessment models)
\item These models are most often parameterised using data for less extreme climates (i.e. temperate or continental climates; and to a lesser extent, boreal and Mediterranean climates. Parameterisation of these models for more extreme bio-geographical regions such as sub arctic climates has not been perfromed and relies on assuming that boreal parameterisations will be applicable.
\item It is important to understand the implications of using these more generic parameterisations for more extreme climate regions, here we explore how risk estimates might be over or underestimated using generic vs bespoke sub-arctic parameterisations
\item we also explore the role [O3] profile and changing climate variables to understand how this risk might alter in the future with the advent of climate change 
\item This allows both an understanding of how the threat from ozone may change over time but also which are the key aspects of plant physiology that might determine risk (and hence which parameters might require special attention to get right) both for pollution impact assessment, but also more generally for describing gas exchange response to changing climates that will influence bio-geochemcial cycling (could we add something in here about the paucity of PFTs representing subarctic conditions in biogeochemcial models?).
\end{itemize}

Ground level ozone is a highly toxic pollutant known to cause damage to a variety of ecosystems around the World \citep{PT:Emberson2020}. At continuously high concentration, ozone negatively affects life expectancy of humans \citep{WHO2008} and is a cause of visible injury, phtosynthetic damage, early senescence as well as programmed cell death \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone related to enhanced pollutant emissions from the transport, residential, power generation and industrial sectors have increased in the past few decades. At the same time, a successful implementation of air quality regulations coupled with economic restructuring which saw a move from coal to gas, reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}.

With a radiative forcing value of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, ozone is a potent climate forcer, ranked third behind \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}. Its damaging effects on the carbon sequestered by global vegetation might provide an important additional feedback to climate change \citep{Nat:Sitch2007}.

In the presence of nitrogen oxides (\chem{NO_x}), tropospheric ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic substances (VOCs). In the absence of local pollution sources, episodes of high ozone concentrations are often attributed to long-range transport of precursors. In northern Fennoscandia, ozone episodes in 2003 and 2006 have been traced back to ozone precursor emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Hydrocarbons are also emitted by living vegetation in form of terpenes and monoterpenes, so called biogenic volatile substances (BVOCs). These emissions are associated with, e.g., thermal stress, defense against herbivores, or attraction of pollinators \citep{TPS:Penuelas2003}.

Despite a relatively short tropospheric life-time of approximately $22\,\unit{days}$, ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone is subject to advection over long distances.
In the annual cycle in boreal and temperate climates, a distinct maximum typically occurs in spring followed by lower concentrations throughout the summer. This is most likely due to stronger dry deposition on land vegetation than on bare ground or snow and ice \citep{AE:Monks2000} causing a build off of ozone and its precursors in the course of polar night. Come spring they are photochemical activated again. This spring peak is reached earlier in northern Fennoscandia than in more southern locations but is highly variable throughout the years \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is a complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

At the same time, the start of growing season in northern Fennoscandia is shifting to earlier days of the year in response to the general warming trend \citep[e.g.,]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby converging with the period of the ozone spring peak. This may promote a higher susceptibility of deciduous species to ozone and could also affect coniferous species which will have flushes of new needles in the spring time which are likely, through their immaturity, to be more susceptible to ozone damage.

Even though long-term (pre 1950) observations of surface ozone concentrations are scarce; evidence from trend analysis of those few sites that have monitored concentrations suggests that since the industrial revolution, background concentrations in the northern Hemisphere have at least doubled and continued to increase until relatively recently \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level-off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}.\\

Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange allowing the entry of CO2 for photosynthesis and loss of H2O vapour via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the course of the day and growing season primarily to balance CO2 uptake against H2O vapour  loss.  The higher the stomatal conductance, the higher the potential for ozone uptake.\\

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for planetary boundary layer ozone formation.

An unusually weak and northward shifted jet stream allowed for a persistent high pressure system above northern Europe, including Fennoscandia which blocked the pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.

Heat and drought stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high pressure system extended even over a period of five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and \chem{VOCs} ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}. Coincident peak \chem{[O_3]} are found in ozone monitoring data from Svanhovd but also over whole northern Fennoscandia in July (see Fig.~\ref{fig:data_svanvik_2018_2019} or Fig.~\ref{fig:ozone_reconstruction_2018_07}).

Like in the 2003 drought period, elevated ozone in Europe was likely promoted by a combination of various other factors as well, such as biogenic VOC emissions, reduced cloud and hence increased solar radiation, faster photochemical reactions, and reduced dry deposition and turbulent mixing due to stagnant weather conditions \citep{JGR:Solberg2018}.\\

To measure the risk of damage inflicted to vegetation through ozone, many different metrics have been defined. Two main types can be differentiated, concentration-based or flux-based metrics, e.g., Accumulated Over a Threshold $\mathrm{x}$ ($\mathrm{AOT_x}$) and Phyto-toxic Ozone Dose over a threshold $\mathrm{y}$ ($\mathrm{POD_y}$), respectively. Due to their simplicity, concentration-based metrics are still widely used, despite the actual integrated flux of ozone entering through the stomata being more relevant for an estimation of the phyto-toxic ozone dose \citep{GCB:Mills2010,ICP:MappingManual2017}.

The $\mathrm{AOT_x}$ in units of $\unit{ppb}$ is defined as:
\begin{equation}
  \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \label{eq:aot}
\end{equation}
with the ambient ozone concentration $\chem{[O_3]}$. The  time over which the integration is calculated varies, but is usually defined from $8\,\unit{am}-8\,\unit{pm}$.
The $\mathrm{POD_y}$ is calculated by first estimating the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf. The terms $r_c$ (leaf surface resistance) and $r_b$ (quasi-laminar resistance) allow for the deposition of \chem{O_3} to the leaf and the fraction that is taken up via the stomata. 
%
\begin{equation}
  \Phi_\mathrm{sto} = \chem{[O_3]} \cdot \frac{u(z_1) \cdot g_\mathrm{sto} \cdot r_c}{r_b + r_c}.
  \label{eq:flux_stomata}
\end{equation}
%
The quasi-laminar boundary layer resistance is calculated by
%
\begin{equation}
  r_b = 1.3 \cdot 150 \cdot \sqrt{\frac{L}{u(z_1)}}.
  \label{eq:quasi_laminar_resist}
\end{equation}
%
Where $L$ is the cross-wind leaf dimension, $u(z_1)$ the wind speed at height $z_1$, and the factor $1.3$ accounts for the diffusivities between heat and \chem{O_3}. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged ambient air ozone flux through the stomata Eq.~(\ref{eq:flux_stomata}) and a flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ is representing the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$).

@@

In terms of $\mathrm{AOT_{40}}$, the threshold for protection of vegetation of $3000\,\unit{ppb}$ (integrated over the growing season) given by the EU directive has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019}. {\bf TODO: might be interesting to look at Mills et al 2003 and the different spatial extents of risk estimated when using AOT40 vs PODy. That could be useful to explain why we focus on PODy (rather then AOT40).- Lisa Come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia.}

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with the Norwegian Institute for Air Research (NILU). 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

As previously mentioned, ozone damage is rather driven by the actual uptake of ozone ($\mathrm{POD_y}$) than by the accumulated ambient air concentration above a certain threshold ($\mathrm{AOT_x}$). Through the stomatal conductance, $\mathrm{POD_y}$ depends on environmental drivers such as ambient air temperature, photoactive radiation, soil water availability, and water vapor pressure deficit.

@@

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive climatologies for surface ozone concentration at Svanhovd and northern Fennoscandia based on in situ observations and also compute climatologies of surface temperature, precipitation, and global radiation for Svanhovd. Based on these, we evaluate anomalies of environmental key variables in 2018/19 by statistical means. In addition, the derived surface ozone concentration climatology for northern Fennoscandia will be compared with available ozone reanalysis products. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate main environmental drivers for the observed ozone damage in 2018 (Section~\ref{sec:do3se}). In this context, we biefely introduce the main $\mathrm{DO_3SE}$ model equations and derive bespoke parameterizations for subarctic species. In Section~\ref{sec:conc} finally, we will summarize our results, draw conclusions, and give an outlook on further work.

\newpage
\section{Data}
\label{sec:data}

@@

@@

>>>>

\Author[1]{Stefanie}{Falk}
\Author[2]{Ane Victoria}{Vollsnes}
\Author[2]{Aud E. B.}{Eriksen}
\Author[3]{Lisa}{Emberson}
\Author[3]{Connie}{O'Neill}
\Author[1]{Frode}{Stordal}
\Author[1]{Terje}{Koren Berntsen}

@@

Ground level ozone is a highly toxic pollutant known to cause damage to a variety of ecosystems around the World \citep{PT:Emberson2020}. At continuously high concentration levels, ozone negatively affects life expectancy of humans \citep{WHO2008} and is a cause of visible injury, photosynthetic damage, early senescence as well as programmed cell death of plants \citep{PCE:Kangasjarvi2005}. Annual global yield losses of four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a suggested loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. {\bf TODO: Wealth of evidence demonstrating ozone damage to vegetation (perhaps focus on key evidence from northern climates and key species (e.g. birch, Norway srpuce etc... - this might help to make it clear why we focus on these species with the PODy modelling))}
The crop yield loss studies indicate a threat to food security in rapidly developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone concentrations related to enhanced pollutant emissions from the transport, residential, power generation and industrial sectors have increased in the past few decades. Even though long-term (pre 1950) observations of surface ozone concentrations are scarce; evidence from trend analysis of those few sites suggests that since the industrial revolution, background concentrations in the northern Hemisphere have at least doubled and continued to increase until relatively recently \citep[Chapter 2]{IPCC2013}. Recent trend analyses indicate that a maximum was reached in 2007 \citep{AE:Derwent2018}, after which tropospheric ozone started to level-off or even decline \citep{ESA:Cooper2014, ACP:Wespes2018,ESA:Gaudel2018}. This can be attributed to a successful implementation of air quality regulations coupled with economic restructuring which saw a move from coal to gas, reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. However, there are indications that an increasing frequency of thermal stress accompanying climate change will reduce vegetations' capacity for ozone uptake and therefor scotch these achievments \citep{NCC:Lin2020}.\\

Ground level ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic substances (VOCs) in the presence of nitrogen oxides (\chem{NO_x}). Despite a relatively short tropospheric life-time of approximately $22\,\unit{days}$, ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone is subject to advection over long distances. Episodes of high ozone concentrations in the absence of local pollution sources can therefor often be attributed to long-range transport of both ozone and its precursors. In northern Fennoscandia, episodes of enhanced ozone concentrations in 2003 and 2006 have been traced back to ozone precursor emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. In addition, hydrocarbons are also emitted by vegetation itself in form of terpenes and monoterpenes, so called biogenic volatile substances (BVOCs). These emissions are associated with, e.g., thermal stress, defense against herbivores, or even attraction of pollinators \citep{TPS:Penuelas2003}.

In the annual cycle in boreal and temperate climates, a distinct maximum of ground level ozone concentrations typically occurs in spring followed by lower concentrations throughout the summer. Several conditions unique to the Actic are stimulating the formation of this spring peak. Major pathways of ozone and its precursors' removal from the boundary layer include dry deposition on vegetated surfaces and photochemistry \citep{RG:Clifton2020}. As dry deposition to bare ground or snow and ice covered surfaces was found to be low \citep{ACP:Helmig2007}, both pathways of removal are substantially suppressed during polar night \citep{AE:Monks2000}. This leads to a build up of ozone and its precursors in the course of winter. Come spring precursors are photochemical activated again, while snow cover typically prevails for a longer period of time continuing to suppress ozone dry deposition.
On the other hand, very high ozone concentrations occuring over shorter periodes of time can often be attributed to deep tropopause folding events over the Actic during winter. These events facilitate an intrusion of stratospheric air masses which are enriched in ozone \citep{JGR:Skerlak2015}.
The spring peak is more pronounced in northern Fennoscandia than in more southern locations but shows a  high interannual variability \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is the complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as described above as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability of ozone concentrations in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

The start of growing season in northern Fennoscandia has been shifting to earlier dates in response to the general warming trend \citep[e.g.,]{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007} and thereby converging with the period of the ozone spring peak. At the same time the growing season is also becoming longer. In conjunction with midnight sun conditions (almost $24\,\unit{h}$ photosynthetic activity), a longer growing season, is prolonging the time in which vegetation can accumulate ozone. These factors may promote a higher potential susceptibility of northern vegetation to ozone in the future.

Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapour via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the course of the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapour loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been emperically linked to environmental factors such as air temperature, photosynthetic active radiation, vapor pressure deficit, and soil water potential as well as photosynthesis itself.\\

To measure the risk of damage inflicted to vegetation through ozone, different metrics have been defined. Two main types can be differentiated, concentration-based or flux-based metrics, e.g., Accumulated Over a Threshold $\mathrm{x}$ ($\mathrm{AOT_x}$) and Phyto-toxic Ozone Dose over a threshold $\mathrm{y}$ ($\mathrm{POD_y}$), respectively. Due to their simplicity, concentration-based metrics are still widely used. Despite comparisons of exposure-based methods with previously used concentration based methods (e.g. $\mathrm{AOT_{40}}$) show that $\mathrm{POD_y}$ is better suited to represent the risk to vegetation from ozone, especially in the more climate extreme parts of Europe \citep{GCB:Mills2010,ICP:MappingManual2017}.

The concentration-based metric $\mathrm{AOT_x}$ (in units of $\unit{ppb}$) is defined as:
\begin{equation}
  \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \label{eq:aot}
\end{equation}
with the ambient ozone concentration $\chem{[O_3]}$. The  time over which the integration is calculated varies, but is usually defined from $8\,\unit{am}-8\,\unit{pm}$.

For computing the $\mathrm{POD_y}$, an estimate of the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) is calculated based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf. Details specific to the $\mathrm{DO_3SE}$ model are given in Section~\ref{sec_do3se}. $\mathrm{POD_y}$ is then calculated according to:
%
\begin{equation}
  \mathrm{POD_y} = \int{(\Phi_\mathrm{sto}-y)\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
%
with the hourly averaged ambient air ozone flux through the stomata Eq.~(\ref{eq:flux_stomata}) and a flux threshold $\mathrm{y}$ both given in units of \unit{nmol\,\chem{O_3}\,m^{-2}\,\mathrm{PLA}\,s^{-1}}. The flux threshold $\mathrm{y}$ is representing the detoxification potential of the plant and is typically only exceeded during daylight hours (i.e. when global radiation is above $50\,\unit{W\,m^{-2}}$).

@@

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for planetary boundary layer ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high pressure system above northern Europe, including Fennoscandia which blocked the pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high pressure system extended even over a period of five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and \chem{VOCs} ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}. Coincident peak \chem{[O_3]} are found in ozone monitoring data from Svanhovd but also over whole northern Fennoscandia in July (see Fig.~\ref{fig:data_svanvik_2018_2019} or Fig.~\ref{fig:ozone_reconstruction_2018_07}).
Like in the 2003 drought period, elevated ozone in Europe was likely promoted by a combination of various other factors as well, such as biogenic VOC emissions, reduced cloud and hence increased solar radiation, faster photochemical reactions, and reduced dry deposition and turbulent mixing due to stagnant weather conditions \citep{JGR:Solberg2018}.\\

These methods described above rely on accurate representation of plant physiological response to climate variables (i.e. the species or plant funtional type specific responses used to parameterise risk assessment models) and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as sub-arctic climates has not been perfromed and relies on assuming that less extreme boreal parameterizations will be applicable.
It is important to understand the implications of using these more generic parameterizations for more extreme climate regions, here we explore how risk estimates might be over or underestimated using generic vs bespoke sub-arctic parameterisations. we also explore the role [O3] profile and changing climate variables to understand how this risk might alter in the future with the advent of climate change. This allows both an understanding of how the threat from ozone may change over time but also which are the key aspects of plant physiology that might determine risk (and hence which parameters might require special attention to get right) both for pollution impact assessment, but also more generally for describing gas exchange response to changing climates that will influence bio-geochemcial cycling. This might be one key to solve long-standing issues of coupled earthsystem models (ESM) which generally lack PFTs representing sub-arctic conditions in their landsurface model.

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with the Norwegian Institute for Air Research (NILU). 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019. In terms of $\mathrm{AOT_{40}}$, the threshold for protection of vegetation of $3000\,\unit{ppb}$ (integrated over the growing season) given by the EU directive has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019} but not at Svanhovd.

@@

The observation site at NIBIO Environment Centre Svanhovd and all related data are presented in Section~\ref{sec:data}. In Section~\ref{sec:stats}, we derive climatologies for surface ozone concentration at Svanhovd and northern Fennoscandia based on in situ observations and also compute climatologies of surface temperature, precipitation, and global radiation for Svanhovd. Based on these, we evaluate anomalies of environmental key variables in 2018/19 by statistical means. In addition, the derived surface ozone concentration climatology for northern Fennoscandia will be compared with available ozone reanalysis products. We use the $\mathrm{DO_3SE}$ model to estimate the ozone uptake by natural and semi-natural vegetation to investigate main environmental drivers for the observed ozone damage in 2018 (Section~\ref{sec:do3se}). In this context, we introduce the$\mathrm{DO_3SE}$ model equations and derive bespoke parameterizations for sub-arctic species to assess systematic uncertainties arising from the use of default parameterizations. In Section~\ref{sec:conc} finally, we will summarize our results, draw conclusions, and give an outlook on further work.

\section{Data}
\label{sec:data}

@@

To compute $\mathrm{POD_y}$, we estimate the stomatal \chem{O_3} flux ($\Phi_\mathrm{sto}$) based on the assumption that the concentration of \chem{O_3} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf. The terms $r_c$ (leaf surface resistance) and $r_b$ (quasi-laminar resistance) allow for the deposition of \chem{O_3} to the leaf and the fraction that is taken up via the stomata. 
%
\begin{equation}
  \Phi_\mathrm{sto} = \chem{[O_3]} \cdot \frac{u(z_1) \cdot g_\mathrm{sto} \cdot r_c}{r_b + r_c}.
  \label{eq:flux_stomata}
\end{equation}
%
The quasi-laminar boundary layer resistance is calculated by
%
\begin{equation}
  r_b = 1.3 \cdot 150 \cdot \sqrt{\frac{L}{u(z_1)}}.
  \label{eq:quasi_laminar_resist}
\end{equation}
%
Where $L$ is the cross-wind leaf dimension, $u(z_1)$ the wind speed at height $z_1$, and the factor $1.3$ accounts for the diffusivities between heat and \chem{O_3}.

@@

{\bf TODO: might be interesting to look at Mills et al 2003 and the different spatial extents of risk estimated when using AOT40 vs PODy. That could be useful to explain why we focus on PODy (rather then AOT40).- Lisa Come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia.}
